Transport map accelerated adaptive importance sampling, and application to inverse problems arising from multiscale stochastic reaction networks

### 2018-01-24 TPAIS_2018.tex
First draft nearly done
Apart from intro and discussion

\title{Transport map accelerated-PAIS, and application to inverse problems arising from
  multiscale stochastic reaction networks}
\author{Simon Cotter (EPSRC and INI), Yannis Kevrekidis (?), Paul
  Russell (EPSRC)}
\begin{document}
\maketitle
\begin{abstract}
In many applications, inverse problems arise where where there are
complex correlations between the different parameters which we wish to
infer from data. The correlations often manifest themselves as lower
dimensional manifolds on which the likelihood function is
invariant, or varies very little. This can be due to trying to infer
unobservable parameters, or due to sloppiness in the model which is
being used to describe the data. In such a situation, standard
sampling methods for characterising the posterior distribution which
do not incorporate information about this structure will be highly
inefficient. Moreover, most methods are inherently serial in nature,
and as such are not expoiting the parallelised  nature of modern
computer infrastructure. In this paper, we seek to develop a method to
tackle this problem, using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds.

@@

\section{Parallel Adaptive Importance Sampling}\label{sec:PAIS}
PAIS****CITE**** is a variant of PMC****CITE****, which is a family of methods which are based
on importance sampling. In importance sampling, we attempt to
characterise a target density through sampling from that
density. However, the target density $\pi$ itself is often too complex to sample
from directly, so we instead sample from a
proposal density $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
sampling from a different distribution to $\pi$. Monte Carlo estimates
using a sample of size $N$
of a function $f$ with respect to $\pi$ can then be made through the
formula
\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
  w_kf(\theta^(k)).\]
This method works well when $\pi$ and $\chi$ are close, but can be
excruciatingly slow when they are not. The idea behind PMC methods is
to construct a good proposal distribution, either from the entire
history of the algorithm up to the current point, or to use the
current state of a whole ensemble of $M$ particles in the system.

In PAIS, the proposal distribution $\chi$ is chosen to be the equally
weighted mixture of any choice of MCMC proposal kernel, evaluated at
each of the current particles in the system. If $\theta^{(k)} = [\theta_1^{(k)},
\theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
ensemble, and we wish to use an MCMC proposal density $q(\cdot ;
\cdot, \beta)$ then 
\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}.\]
Often the variance of the MCMC proposal kernels can be tuned using
their respective algorithmic parameters $\beta$. Good values for these
algorithmic parameters can be
found by optimising for the effective sample size of the importance
sample that is produced (see ***CITE*** for more details).

If the ensemble is large enough, and the chain has entered
probabilistic stationarity, then the current state of the ensemble is
a good rough discrete approximation of the target density, and in turn
$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
sample $\{\hat{\theta}^{(k)},w^{(k)}\}$. It can be advantageous to use stratified sampling of the
mixture in order to
ensure that the sample made is as representative as possible of the
target density, i.e.
\[ \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)}\]
for each $i = 1,2,\ldots,M$. We now have a weighted sample, and it
would be inadvisable to use an equally weighted mixture of proposal
distributions from each of these points. Therefore, before starting
the next iteration, the importance sample
$\{\hat{\theta}^{(k)},w^{(k)}\}$ is resampled to produce an equally
weighted sample, ready for the next iteration of the algorithm. In
PAIS, a state-of-the-art resampler is used, which uses optimal
transport methods to find the equally weighted discrete sample which
best represents the statistics of the importance
sample***CITE****. For larger ensemble sizes $M$ this can become
expensive, in which case a greedy approximation of this algorithm, the
Approximate Multinomial Resampler (AMR) can be
implemented***CITE****. The output of the resampler is then denoted
$\theta^{(k+1)}$ and the algorithm is ready for the next
iteration. The importance samples $\{\hat{\theta}^{(k)},w^{(k)}\}$ The PAIS is summarised in Algorithm \ref{alg:PAIS}.

@@

One problem with the PAIS and other PMC methods can become apparent if
the target density has very strong correlations in its structure, in
particular if that correlation is not global but only local. In this
case, unless the proposal densities $q$ are informed by this local
structure, the mixture distribution proposal may not well approximate
$\pi$ without a very large large ensemble size $M$, which can become
inhibitively expensive. Some methods have been proposed ****CITE****
which use samples local to each particle to inform local covariance
structure.

@@

\begin{dfn}[(Exact) Transport Map $T$]
	A transport map $T$ is a function $T\colon
        \mathcal{X}\rightarrow\mathbb{R}^d$ such that the {\it
          pullback} of the reference measure with density $\phi(\cdot)$,
	\begin{equation}\label{eq:pullback}
		\tilde{\pi}(\theta) = \phi(T(\theta))|J_T(\theta)|,
	\end{equation}
	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
	\[
		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
	\]
\end{dfn}

@@

\begin{figure}[htpb]
\centering
\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
\caption{Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
\label{fig:R2_transport}
\end{figure}

@@

We first find the optimal scaling parameters for the individual
algorithms. This is done, as in ****CITE US**** by optimising for the effective sample size in
the PAIS algorithm, and by tuning the relative $L^2$ error in the MH
algorithm. There is currently no
guidance on the best way of tuning the MH algorithm with transport map
proposals although one might expect results similar to the standard MH
results, especially if adaptation of the map is stopped after a given
point. As in the PAIS algorithm, optimising for the effective sample size might be the best option.

@@

\begin{figure}[!ht]
\centering
\subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
\subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
\caption{Convergence of Algorithms Transport M-H,
  \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
\label{fig:R2_l2_convergence}
\end{figure}

@@

The model can be represented as the following expression involving
$N_r$ different unit rate Poisson
processes****CITE ANDERSON ET AL**** $Y_j$, given by:
\begin{equation}\label{eq:RTC}
X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
\end{equation}

The master equation for this model is only solvable in certain
situations, for example monomolecular networks****CITE****, or for
steady-state solutions certain deficiency
zero networks****CITE*****. Trajectories for this system can be
sampled exactly, for instance using the Gillespie SSA***CITE***, or
its variants\cite{gillespie2007stochastic}****CITE****. However, if the system is stiff, i.e. there
are some reactions which are firing many times on a timescale for
which others are unlikely to fire at all, then trajectories can become
prohibitively expensive to simulate, since these methods simulate
every single reaction event with the same cost. In such a system, one
might employ multiscale methods in order to approximate trajectories
at a lower cost than the exact algorithms.

@@

This approach can work well when there is a large timescale gap
between the fast and slow reactions in a system, but where the
timescale gap is less pronounced, it can lead to significant
errors***cite cotter erban errors paper***. Another approach is the
constrained multiscale method (CMA)***CITE***, based in part on the
equation-free approach to multiscale computations***CITE
YANNIS****. This approach also assumes quasi-equillibrium in the
system, but takes into account the effect that the slow reactions can
have on the invariant distribution of the fast variables in the
system. For a more detailed description of this method, please refer
to the literature***CITATIONS***.

@@

First we consider a simple multiscale example involving two chemical
species $S_1$ and $S_2$ involved in only zeroth and first order reactions:
\begin{equation}\label{eqn:full_system}
	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
\end{equation}
Each arrow represents a reaction from a reactant to a product, with
some rate constant $k_i$, and where mass action kinetics is
assumed. The parameters $k_i$ are non-negative, and $\mathbf{k} =
[k_1,\dots,k_4]^\top \in \mathbb{R}_+^4 = \mathcal{X}$. We denote the
population count of
species $S_i$ by $X_i \in \mathbb{N}_0$. We assume that we are in a
parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
chemical species are involved in fast reactions. However, the quantity
$S = X_1 + X_2$, is conserved by both of the fast reactions,
and as such, this is the slowly changing quantity in this system.

We assume that we are only able to observe
the slow variable $S(t)$. 
The effective dynamics of $S$ can be approximated by a system of only
two reactions:
\begin{equation}\label{eqn:QSSA_system}
	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset.
\end{equation}
The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
application of a multiscale approximation, as detailed in Section
\ref{sec:multi}, and in more detail in ****CITE****. These effective
propensities can often be computed without the need for expensive
simulations, in particular when the fast subsystem is deficiency zero,
which is often the case****CITE DIF ZERO PAPERS****. For this very
simple example, the dependance of these effective
propensities on the original rate parameters can be explicitly
understood. Under the QEA,
the effective rate of degradation of the slow variable $S$ is given by
\[
	\bar{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3}.
\]
Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
\begin{equation}\label{eqn:chem_CMA_rate}
	\bar{\alpha}_4^{\text{CMA}}(s) = \mathbb{E}_{\text{CMA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3+k_4}.
\end{equation}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
\caption{Posterior for the constrained multiscale problem outlined in Section~\ref{sec:chem_multiscale}.}
\label{fig:chem_CMA_posterior}
\end{figure}

@@

		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
	\end{tikzpicture}
	\caption{Top: Couplings between parameter space and reference
          when using the transport map. Bottom: Couplings between parameter space and reference space
          when using the transport map proposal with a $\log$ preconditioner.}
	\label{fig:chem_coupling}
\end{figure}

@@

% choice of resampler?
We will use the AMR***CITE*** resampler with an ensemble size of
$M=500$. This increase in ensemble size in comparison with previous
example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.

@@

\begin{figure}[!htb]
\centering
\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
\caption{Convergence of the constrained multiscale example described in Section~\ref{sec:chem_multiscale}.}
\label{fig:chem_multiscale_L2}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\subfigure[Reference space mapped to by $\hat{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
\subfigure[Reference space mapped to by $T\circ\log$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
\caption{Push forward of the target measure on reference space for the problem described in Section~\ref{sec:multiscale_posterior} found using the TRW and logTRW proposal distributions. Components are linked by the relation $r_i = T_i'(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
\label{fig:chem_reference_spaces}
\end{figure}

@@

\begin{table}[!htb]
\centering
\begin{tabular}{rr}
	\toprule
	$k$ & \quad $\mathbb{P}(D = (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$ \\ \cmidrule(lr){1-2}
	0 & 6.8e-3 \\
	1 & 3.2e-3 \\
	2 & 1.7e-3 \\ \bottomrule
\end{tabular}
\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:multiscale_posterior}.}
\label{tab:chem_Bayes_marginals}
\end{table}

@@

\begin{figure}[!htb]
\centering
\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/QEA_approx_marginal"}}
\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
\caption{Comparison of the approximate marginal densities for the `observable parameter' $\hat{k}_4$ under models $\mathcal{M}_1$ and $\mathcal{M}_2$. Marginals for these two parameters are approximated using samples which have been produced by targeting the three densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
\label{fig:chem_model_comp}
\end{figure}

@@

The effective dynamics of $T$ and $M$ are then given by:
\begin{align}
	\emptyset \xrightarrow{\hat{\alpha}_5} M \quad & \quad M\xrightarrow{\alpha_6} M+T  \label{eqn:GRN_eff} \\
	T \xrightarrow{\hat{\alpha}_7} \emptyset \quad & \quad M \xrightarrow{\alpha_8} \emptyset \nonumber
\end{align}
Here, the propensities are displayed, as opposed to the rates, with
approximations of the propensities $\hat{\alpha}_5$ and
$\hat{\alpha}_7$ to be computed. We omit the derivation of these
effective propensities, using the CMA methodology, for brevity, but
interested readers can see more details in ****CITE THESIS****.

@@

\begin{figure}[htb]
\centering
\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
\caption{Marginal densities of the posterior distribution for the GRN example in Section~\ref{sec:grn}.}
\label{fig:GRN_posterior}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
\caption{Convergence of the GRN example described in Section~\ref{sec:grn}.}
\label{fig:grn_L2}
\end{figure}

@@

\subsection{Sampling in (moderately) higher
  dimensions}\label{sec:TPAIS_higher_dim}
One major problem with importance sampling schemes is the curse of
dimensionality, which means that methods such as PAIS, and other PMC
methods, can only be used for relatively low dimensional
problems. Here, we will briefly discuss how transport maps could aid
with making moderately higher dimensional problems accessible to this
family of methods.

>>>>

\author{Simon L. Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1. SLC would like to thank the Isaac Newton
    Institute for Mathematical Sciences for support and hospitality
    during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
EPSRC grant number EP/K032208/1.} \and Ioannis G. Kevrekidis\thanks{} \and Paul
  Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK.}}

\title{Transport map accelerated adaptive importance sampling, and application to inverse problems arising from
  multiscale stochastic reaction networks}
\begin{document}
\maketitle
\begin{abstract}
In many applications, inverse problems arise where where there are
complex correlations between the different parameters which we wish to
infer from data. The correlations often manifest themselves as lower
dimensional manifolds on which the likelihood function is
invariant, or varies very little. This can be due to trying to infer
unobservable parameters, or due to sloppiness in the model which is
being used to describe the data. In such a situation, standard
sampling methods for characterising the posterior distribution which
do not incorporate information about this structure will be highly
inefficient. Moreover, most methods are inherently serial in nature,
and as such are not expoiting the parallelised  nature of modern
computer infrastructure. In this paper, we seek to develop a method to
tackle this problem, using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds.

@@

\section{Parallel Adaptive Importance Sampling}\label{sec:PAIS}
Parallel adaptive importance sampling (PAIS)\cite{cotter2015parallel} is a variant of PMC\cite{cappe2012population}, which is a family of methods which are based
on importance sampling. In importance sampling, we attempt to
characterise a target density through sampling from that
density. However, the target density $\pi$ itself is often too complex to sample
from directly, so we instead sample from a
proposal density $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
sampling from a different distribution to $\pi$. Monte Carlo estimates
using a sample of size $N$
of a function $f$ with respect to $\pi$ can then be made through the
formula
\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
  w_kf(\theta^(k)).\]
This method works well when $\pi$ and $\chi$ are close, but can be
excruciatingly slow when they are not. The idea behind PMC methods is
to construct a good proposal distribution, either from the entire
history of the algorithm up to the current point, or to use the
current state of a whole ensemble of $M$ particles in the system.

In PAIS, the proposal distribution $\chi$ is chosen to be the equally
weighted mixture of any choice of MCMC proposal kernel, evaluated at
each of the current particles in the system. If $\theta^{(k)} = [\theta_1^{(k)},
\theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
ensemble, and we wish to use an MCMC proposal density $q(\cdot ;
\cdot, \beta)$ then 
\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}.\]
Often the variance of the MCMC proposal kernels can be tuned using
their respective algorithmic parameters $\beta$. Good values for these
algorithmic parameters can be
found by optimising for the effective sample size of the importance
sample that is produced (see \cite{cotter2015parallel, russ2017parallel} for more details).

If the ensemble is large enough, and the chain has entered
probabilistic stationarity, then the current state of the ensemble is
a good rough discrete approximation of the target density, and in turn
$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
sample $\{\hat{\theta}^{(k)},w^{(k)}\}$. It can be advantageous to use stratified sampling of the
mixture in order to
ensure that the sample made is as representative as possible of the
target density, i.e.
\[ \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)}\]
for each $i = 1,2,\ldots,M$. We now have a weighted sample, and it
would be inadvisable to use an equally weighted mixture of proposal
distributions from each of these points. Therefore, before starting
the next iteration, the importance sample
$\{\hat{\theta}^{(k)},w^{(k)}\}$ is resampled to produce an equally
weighted sample, ready for the next iteration of the algorithm. In
PAIS, a state-of-the-art resampler is used, which uses optimal
transport methods to find the equally weighted discrete sample which
best represents the statistics of the importance
sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
expensive, in which case a greedy approximation of this algorithm, the
Approximate Multinomial Resampler (AMR) can be
implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
$\theta^{(k+1)}$ and the algorithm is ready for the next
iteration. The importance samples $\{\hat{\theta}^{(k)},w^{(k)}\}$ The PAIS is summarised in Algorithm \ref{alg:PAIS}.

@@

One problem with the PAIS and other PMC methods can become apparent if
the target density has very strong correlations in its structure, in
particular if that correlation is not global but only local. In this
case, unless the proposal densities $q$ are informed by this local
structure, the mixture distribution proposal may not well approximate
$\pi$ without a very large large ensemble size $M$, which can become
inhibitively expensive. Some methods have been
proposed\cite{douc2007minimum}
which use samples local to each particle to inform local covariance
structure.

@@

\begin{dfn}[Transport Map $T$]
	A transport map $T$ is a function $T\colon
        \mathcal{X}\rightarrow\mathbb{R}^d$ such that the {\it
          pullback} of the reference measure with density $\phi(\cdot)$,
	\begin{equation}\label{eq:pullback}
		\tilde{\pi}(\theta) = \phi(T(\theta))|J_T(\theta)|,
	\end{equation}
	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
	\[
		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
	\]
\end{dfn}

@@

\begin{figure}[htpb]
\centering
\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
\caption{The effect of the approximate transport map $\tilde{T}$ on a
  sample from the Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
\label{fig:R2_transport}
\end{figure}

@@

We first find the optimal scaling parameters for the individual
algorithms. This is done, as in \cite{cotter2015parallel}, by optimising for the effective sample size in
the PAIS algorithm, and by tuning the relative $L^2$ error in the MH
algorithm. There is currently no
guidance on the best way of tuning the MH algorithm with transport map
proposals although one might expect results similar to the standard MH
results, especially if adaptation of the map is stopped after a given
point. As in the PAIS algorithm, optimising for the effective sample size might be the best option.

@@

\begin{figure}[!ht]
\centering
\subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
\subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
\caption{Convergence of algorithms; Transport M-H,
  algorithm \ref{alg:TransportPAIS1}, algorithm \ref{alg:TransportPAIS2} for
  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
\label{fig:R2_l2_convergence}
\end{figure}

@@

The model can be represented as the following expression involving
$N_r$ different unit rate Poisson
processes\cite{anderson2011continuous} $Y_j$, given by:
\begin{equation}\label{eq:RTC}
X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
\end{equation}

The master equation for this model is only solvable in certain
situations, for example monomolecular networks\cite{jahnke2007solving}, or for
steady-state solutions certain deficiency
zero networks\cite{anderson2010product,anderson2016product}. Trajectories for this system can be
sampled exactly, for instance using the Gillespie SSA\cite{gillespie1977exact}, or
its variants\cite{gillespie2007stochastic,gibson1998efficient,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
are some reactions which are firing many times on a timescale for
which others are unlikely to fire at all, then trajectories can become
prohibitively expensive to simulate, since these methods simulate
every single reaction event with the same cost. In such a system, one
might employ multiscale methods in order to approximate trajectories
at a lower cost than the exact algorithms.

@@

This approach can work well when there is a large timescale gap
between the fast and slow reactions in a system, but where the
timescale gap is less pronounced, it can lead to significant
errors\cite{cotter2016error}. Another approach is the
constrained multiscale method (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equillibrium in the
system, but takes into account the effect that the slow reactions can
have on the invariant distribution of the fast variables in the
system. For a more detailed description of this method, please refer
to the literature\cite{cotter2016constrained}.

@@

First we consider a simple multiscale example involving two chemical
species $S_1$ and $S_2$ involved in only zeroth and first order reactions:
\begin{equation}\label{eqn:full_system}
	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
\end{equation}
Each arrow represents a reaction from a reactant to a product, with
some rate constant $k_i$, and where mass action kinetics is
assumed. The parameters $k_i$ are non-negative, and $\mathbf{k} =
[k_1,\dots,k_4]^\top \in \mathbb{R}_+^4 = \mathcal{X}$. We denote the
population count of
species $S_i$ by $X_i \in \mathbb{N}_0$. 
We consider this system with the following parameters:
\begin{equation}\label{eq:params1}
k_1 = 100, \qquad k_2 = k_3 = 10, \qquad k_4 = 1,
\end{equation}

 In this 
parameter regime the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
more frequently than the other reactions. Notice that both
chemical species are involved in fast reactions. However, the quantity
$S = X_1 + X_2$, is conserved by both of the fast reactions,
and as such, this is the slowly changing quantity in this system.

We observe this system over the time period $t \in [0,500]$ with
initial condition ******, but we assume that we are only able to observe
the slow variable $S(t)$. 
The effective dynamics of $S$ can be approximated by a system of only
two reactions:
\begin{equation}\label{eqn:QSSA_system}
	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset,
\end{equation}
where here the propensities are shown as opposed to the rates.

The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
application of a multiscale approximation, as detailed in Section
\ref{sec:multi}, and in more detail in \cite{cotter2016constrained}. These effective
propensities can often be computed without the need for expensive
simulations, in particular when the fast subsystem is deficiency zero,
which is often the case\cite{anderson2010product,anderson2016product}. For this very
simple example, the dependance of these effective
propensities on the original rate parameters can be explicitly
understood. Under the QEA,
the effective rate of degradation of the slow variable $S$ is given by
\[
	\bar{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3}.
\]
Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
\begin{equation}\label{eqn:chem_CMA_rate}
	\bar{\alpha}_4^{\text{CMA}}(s) = \mathbb{E}_{\text{CMA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3+k_4}.
\end{equation}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
\caption{CMA approximation \eqref{eq:Apos} of the posterior arising from observations of
  the slow variable $S = X_1 + X_2$ for system
  \eqref{eqn:full_system}  with true rates given by \eqref{eq:params1}.}
\label{fig:chem_CMA_posterior}
\end{figure}

@@

		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
	\end{tikzpicture}
	\caption{Couplings between parameter space and reference
          when using the transport map with (bottom) and without (top)
        a log preconditioner.}
	\label{fig:chem_coupling}
\end{figure}

@@

% choice of resampler?
We will use the AMR\cite{russ2017parallel} resampler with an ensemble size of
$M=500$. This increase in ensemble size in comparison with previous
example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.

@@

\begin{figure}[!htb]
\centering
\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
\caption{Convergence of eight different MCMC algorithms for the CMA
  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}.}
\label{fig:chem_multiscale_L2}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\subfigure[Push forward of the posterior sample through$\hat{T}$ with no log preconditioner.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
\subfigure[Push forward of the posterior sample through$\hat{T}$ with log preconditioner.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
\caption{Push forward of the target measure on reference space for the the CMA
  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}, found using the TRW and logTRW proposal distributions. Components are linked by the relation $r_i = T_i(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
\label{fig:chem_reference_spaces}
\end{figure}

@@

\begin{table}[!htb]
\centering
\begin{tabular}{rr}
	\toprule
	$k$ & \quad $\mathbb{P}(D = (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$ \\ \cmidrule(lr){1-2}
	0 & 6.8e-3 \\
	1 & 3.2e-3 \\
	2 & 1.7e-3 \\ \bottomrule
\end{tabular}
\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:chem_multiscale}.}
\label{tab:chem_Bayes_marginals}
\end{table}

@@

\begin{figure}[!htb]
\centering
\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/QEA_approx_marginal"}}
\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
\caption{Comparison of the approximate marginal densities for the
  quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$
  across the three posterior densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
\label{fig:chem_model_comp}
\end{figure}

@@

The effective dynamics of $T$ and $M$ are then given by:
\begin{align}
	\emptyset \xrightarrow{\hat{\alpha}_5} M \quad & \quad M\xrightarrow{\alpha_6} M+T  \label{eqn:GRN_eff} \\
	T \xrightarrow{\hat{\alpha}_7} \emptyset \quad & \quad M \xrightarrow{\alpha_8} \emptyset \nonumber
\end{align}
Here, the propensities are displayed, as opposed to the rates, with
approximations of the propensities $\hat{\alpha}_5$ and
$\hat{\alpha}_7$ to be computed. We omit the derivation of these
effective propensities, using the CMA methodology, for brevity, but
interested readers can see more details in \cite{russ2017parallel}.

@@

\begin{figure}[htb]
\centering
\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
\caption{Marginal densities of the CMA approximation of the posterior
  distribution for the GRN example in Section~\ref{sec:grn}.}
\label{fig:GRN_posterior}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
\caption{Convergence of eight different MCMC algorithms for the CMA
  approximation of the  posterior arising from the example in Section \ref{sec:grn}.}
\label{fig:grn_L2}
\end{figure}

@@

One major issue with PMC-type algorithms is the curse of
dimensionality. This curse is two-fold, since as the target
distribution's dimension is increased, a larger ensemble size is often
required in order to capture the addition complexity of the density
that the higher dimension enables. Furthermore, the cost of the
resampling step within the algorithm grows with the ensemble size. For
example, the ETPF algorithm is $\mathcal{O}(n^2)$. However, the
simiplification that the transport map makes has some potential to go
some way to tackling this problem.

\subsection{Sampling in (moderately) higher
  dimensions}\label{sec:TPAIS_higher_dim}
Here, we will briefly discuss how transport maps could aid
with making moderately higher dimensional problems accessible to this
family of methods.
